Asymmetric tail-dependence in extreme climatic events affects population-extinction risk

### 2019-08-27 MainText.Rmd
editing methods...continued

<!--our objective ?-->
In this paper, we will explore the potential use of asymmetric tail-dependence in a copula 
for understanding the extinction risk of populations towards extreme climatic event
by addressing the following questions: (Q1) Does nature of tail dependence 
(left tail or right tail) in vital rates
of a single-patch age-structured population model affect extinction risk of that population differently?
We hypothesized population might have higher extinction risk if vital rates are left-tail dependent.
(Q2) How does asymmetric noise copula (tail dependence across multiple patches) influence extinction 
risk of a density-dependent metapopulation model? Our speculation is that the general finding 
from density-independent metapopulation model [@Ghosh_copula]
(higher extinction risk if noises across patches were 
left-tail dependent than the one if noises had right-tail dependence) would hold true only in
the under-compensatory regime, but should be reverse if the model shows over-compensatory dynamics. 
(Q3) We further hypothesize that when there is no dispersal, the above-said switching should occur
near the critical parameter value where the deterministic model shows transition from 
under-compensatory to over-compensatory dynamics.

@@

<!--stochastic matrix model : age structured-->
To investigate (Q1) from \nameref{Introduction} that how copula structure in noisy 
environment affect the extinction risk of
populations in a single patch age-structured model,
we considered a basic stochastic age-structured model (Eq. \ref{eq_age_str}) 
with two age classes (egg $E$ and female adult $A$). We assumed
the population was semelperous, i.e., adult ($A$) went through the
reproductive episode only once before death. Eggs and adults in the 
next generation depend upon the fecundity of adult ($f_a \geq 0$) 
and survival of eggs ($0 \leq s_e \leq 1$) from present generation, respectively.
The environmental stochasticity in the model was embedded in 
the fluctuating vital rates ($f_a$ and $s_e$). We considered the fecundity ($f_a$)
and survival ($s_e$) rates were random draw from a gamma ($\Gamma(k,\theta)$) and
a beta ($\mathcal{B}(\alpha,\beta)$) distribution, respectively. These distributions 
were generated from iid noises showing specific tail-dependence (either in left- or in right-tail)
between fecundity and survival rates. Tail dependence emerged if extreme bad years ("catastrophs") 
would suppress both vital rates (hence, $f_a$ and $s_e$ would show a left tail dependent copula), whereas 
extreme good years ("bonanzas") would increase both of them (i.e., $f_a$ and $s_e$ would show 
a right tail dependent copula). Eq. \ref{eq_vitalrates} represent the stochasticity 
in the vital rates. We included the density dependence effect in one of the vital rate,
namely, fecundity ($f_a$). If $\Omega = \infty$ in Eq. \ref{eq_vitalrates}, 
the projection matrix would be independent
of populations from any age-class and the model was then called to be "linear", whereas
its "nonlinear" version had finite $\Omega$.

@@

<!--calculation of ext risk with age-structured model-->
To calculate extinction risk after each time step, we simulated model (\ref{eq_age_str}) for $25$ years
with $10000$ simulations. For each simulation, we started with initial population $(E,A)_{t=0} = (5,10)$
and following Eq. \ref{eq_age_str} we checked whether after each time step the total population $(E+A)_{t}$
was below a certain extinction threshold (we set it as $2$) or not. If it was below that threshold then we 
considered population went extinct for that simulation. We also calculated $2.5^{th}$ and $97.5^{th}$ quantiles
of $10000$ simulations with the probability of computed extinction risk at each time step.

To test our hypothesis from (Q1), we repeated the above analysis 
considering the vital rates coming from an asymmetric copula with specific 
left (Clayton, Survival Gumbel)
or right (Survival Clayton, Gumbel) tail dependence
or from an extreme tail-depentdent one, 
when vital rates were considered to be affected
by extreme climatic events. A $N \times 2$ matrix, M, was generated
using `BiCopSim` function from `VineCopula` package 
for each of the left (Clayton, Survival Gumbel) and right 
(Survival Clayton, Gumbel) tail-dependent bivariate copulas, 
where $N$ is number of years ($25$) multiplied by number of simulations ($10000$) we used. 
We kept the same Kendall's correlation ($\tau = 0.6$) for 
each bivariate copula from one parameter family which used its parameter accordingly
(see `iTau` function from `copula` package). Entries within each clumn of
matrix $M$, were thus 
independent of each others but 
the two columns of $M$ had a specified Kendall's 
correlation ($\tau = 0.6$) and underlying tail-dependence. 
Similarly, for extreme-tail dependent bivariant copula
(either left or right tail) a matrix M was generated from standard iid as described in 
Appendix \ref{SM-extr_tail_dep}.

We made the first column of matrix 
$M$ as gamma ($\Gamma(k,\theta)$) distributed. 
$k = 2$ and $\theta = 1$ were shape and scale parameters, 
respectively. This column represented the distribution of fecundity of the adult population, $f_a$.
For a given $\theta$, increasing 
$k$ indicates distribution will shift towards higher fecundity which might 
reduce the extinction risk if the other parameters
remain unchanged. We took the second column of $M$ and applied a 
beta ($\mathcal{B}(\alpha,\beta)$) transformation. $\alpha =3$ and
$\beta = 2$ were the shape parameters which could control the distribution 
of the survival rates of eggs, $s_e$. 
For each $25$ timesteps from each of $10000$ simulations,
the vital rates (fecundity $f_a$ and survival rates $s_e$) 
of the stochastic projrction matrix in model (\ref{eq_age_str})
were drawn pairwise from each row of matrix $M$.
[`sim_age_str`](https://github.com/sghosh89/ERC/blob/master/ext_risk_cop_age_str.R) 
is the function we used to calculate extinction risk for age structured model.

<!--density-dependence metapopulation model-->
We explored (Q2) by examining how nature of the tail dependence of an
environmental variable across space could influence the extinction risk of a 
metapopulation model with density-dependence feedback. We considered six density-dependent 
population-growth models (Ricker, Hassell, Maynard Smith, Pennycuick, Verhulst, Austin-Brewer)
in noisy environment (see Table \ref{tab_model_summary}) 
from a metapopulation context.
Models were adapted from earlier work
by @cohen1995unexpected, excluding only two : Varley model (this is not a continuous one) 
and Malthus-Condorcet-Mill model (terms containing logarithm of extinct population does not make sense). 
All the models could be represented via 
a generalized form : $\vec{P}_{t+1} = D \lambda_t \vec{P_t}$,
where $\vec{P_t}$ was a vector of population densities, 
the $i^{th}$ component $P_t^i$ representing
the density in the $i^{th}$ habitat patch ($i=1,2,\ldots,n$) at time $t$. 
The $n \times n$ matrix $\lambda(t)$ was diagonal with $i^{th}$ diagonal entry 
of stochastic density dependence form $\lambda_i(t)$ (see second column 
of Table \ref{tab_model_summary}). For simplicity, 
we considered $n$ patches in each model were arranged, evenly spaced, in a line
with the provision of "local" or "global" dispersal among patches. A n $\times$ n dispersal
matrix, $D$ was constructed assuming that a fraction $d$ of 
each population patch dispersed during any given time step, equally distributed to
the two or one nearest neighbors in case of local dispersal. For global dispersal, the fraction
$d$ that dispersed
from each patch was divided up equally among all other patches. We called $d$ as dispersal
rate with $d$ being zero there was no dispersal, and threby no chance to rescue an extinct 
patch through migration from nearby metapopulation patches.

The environmental noises from any two out of $n$ patches, say, 
$(\epsilon_t^i, \epsilon_t^j)$ were normally 
distributed within each patch with mean $m=0$ and standard deviation $\sigma$ 
(see the specified value $\sigma$ for each model in 3$^{rd}$ 
column of Table \ref{tab_model_summary}). Noises were 
independent through time, and showed the same 
spatial correlations between patches (to within sampling error) for every simulation, 
but they were made to exhibit extreme right- or extreme left-tail dependence between patches in
different simulations. A three dimensional array, numsims $\times$ n $\times$ numsteps, 
$\epsilon$, was made from these extreme tail-dependence noise 
matrix generated via algorithm stated in 
Appendix \ref{SM-extr_tail_dep} and used to calculate extinction risk 
for each of the specified model in Table \ref{tab_model_summary} for $numsteps = 25$ years, $n =$ 5 patches
and $numsims = 10000$ simulations.

>>>>

<!--our objective ?-->
In this paper, we will explore the potential use of asymmetric tail-dependence in a copula 
for understanding the extinction risk of populations towards extreme climatic event
by addressing the following questions: (Q1) Does nature of tail dependence 
(left tail or right tail) in vital rates
of a single-patch age-structured population model affect extinction risk of that population differently?
We hypothesized population might have higher extinction risk if vital rates are left-tail dependent.
(Q2) How does environmental noises with asymmetric tail dependence across multiple patches
influence extinction risk of a density-dependent metapopulation model? Our speculation 
is that the general finding from density-independent metapopulation model [@Ghosh_copula]
(higher extinction risk if noises across patches were left-tail dependent than the one 
if noises had right-tail dependence) would hold true only in the under-compensatory regime, but 
should be reverse if the model shows over-compensatory dynamics. 
(Q3) We further hypothesize that when there is no dispersal, the above-said switching should occur
near the critical parameter value where the deterministic model shows transition from 
under-compensatory to over-compensatory dynamics.

@@

<!--stochastic matrix model : age structured-->
To investigate (Q1) from \nameref{Introduction} that how does asymmetric tail 
dependent noise affect the extinction risk of
populations in a single patch age-structured model,
we considered a stochastic age-structured model (Eq. \ref{eq_age_str}) 
with two age classes (egg $E$ and female adult $A$). We assumed
the population was semelperous, i.e., adult ($A$) went through the
reproductive episode only once before death. Eggs and adults in the 
next generation evolved depending on the fecundity of adult ($f_a \geq 0$) 
and survival of eggs ($0 \leq s_e \leq 1$) from present generation.
Fluctuating vital rates ($f_a$ and $s_e$) captured the environmental
stochasticity (Eq. \ref{eq_vitalrates}). We made adult's fecundity 
($f_a$) negative density-dependent with its own population. When $\Omega = \infty$ in 
Eq. \ref{eq_vitalrates}, the projection matrix was independent of populations 
(i.e., Eq. \ref{eq_age_str} was "linear" model then), 
but when $\Omega$ was finite, Eq. \ref{eq_age_str} became "nonlinear".
Fecundity ($f_a$) and survival ($s_e$) rates were random draw from a gamma ($\Gamma(k,\theta)$) and
a beta ($\mathcal{B}(\alpha,\beta)$) distribution, respectively. We generated those distributions 
from iid noises with a varied degree of asymmetric tail-dependence between 
fecundity ($f_a$) and survival ($s_e$) rates. Degree of tail dependence had two broad 
categories: moderate and extreme. Each category had specific direction in tail dependence: 
either in left or in right tail of vital rates. Vital rates ($f_a$ and $s_e$) were
left tail dependent when unusual bad years ("catastrophs") suppressed both vital rates synchronously, 
and right tail dependent when unexpected good years ("bonanzas") boosted both of them.

@@

We used bivariate Clayton copula to model moderate left-tail and Survival Clayton copula to model 
moderate right tail dependence in the vital rates. Both copula had the same Kendall 
correlation ($=0.6$) but opposite dependence in their respective tails. 
We generated a $N \times 2$ matrix, M, for each of the left (Clayton) and 
right (Survival Clayton) tail-dependent bivariate copulas (see `BiCopSim` function from
`VineCopula` package ), where $N$ is number of years ($=25$) multiplied 
by number of simulations ($=10000$) we used. 
We chose parameters for each copula family using 
`iTau` function from `copula` package so that both copula had
same Kendall's correlation ($\tau = 0.6$). Entries within each clumn of
matrix $M$, then became independent of each others but 
entries between columns had a specific tail dependence
with positive Kendall's correlation ($\tau = 0.6$). 
Similarly, we generated extreme left- (or right-) tail dependent matrix M 
from standard iid (Appendix \ref{SM-extr_tail_dep}) and made the marginals uniform so that
it could mimic an extreme tail dependent copula.

We made the first column of matrix 
$M$ as gamma ($\Gamma(k,\theta)$) distributed 
with shape parameter $k = 2$ and scale parameter $\theta = 1$.
This distribution represented fecundity of the adult population, $f_a \geq 0$.
For a given $\theta$, increasing $k$ would shift the distribution towards
higher fecundity and reduce the extinction risk if the other parameters
remain unchanged. We took the second column of $M$ and applied a 
beta ($\mathcal{B}(\alpha,\beta)$) transformation. $\alpha =3$ and
$\beta = 2$ were the shape parameters that could control the distribution 
of the survival rates of eggs ($s_e$) ranging in between 0 to 1. 
For each $25$ timesteps from each of $10000$ simulations,
we feed pairwise entries from each row of matrix $M$ as the vital rates 
(fecundity $f_a$ and survival rates $s_e$) 
of the stochastic projrction matrix in model (\ref{eq_age_str}).

<!--calculation of ext risk with age-structured model-->
To calculate extinction risk after each time step, we simulated model (\ref{eq_age_str}) for $25$ years
with $10000$ simulations. For each simulation, we started with initial population $(E,A)_{t=0} = (5,10)$
and following Eq. \ref{eq_age_str} we checked whether after each time step the total population $(E+A)_{t}$
was below a certain extinction threshold (we set it as $2$) or not. If total population was below that threshold then we 
considered population went extinct for that simulation. We also calculated $2.5^{th}$ and $97.5^{th}$ quantiles
of $10000$ simulations with the probability of computed extinction risk at each time step (for details see [`code`](https://github.com/sghosh89/ERC/blob/master/ext_risk_cop_age_str.R)).

<!--density-dependence metapopulation model-->
We explored (Q2) by examining how nature of the tail dependence of an
environmental variable across space could influence the extinction risk of a 
metapopulation model with density-dependence feedback. We considered six density-dependent 
population-growth models (Ricker, Hassell, Maynard Smith, Pennycuick, Verhulst, Austin-Brewer)
in noisy environment (see Table \ref{tab_model_summary}) 
from a metapopulation context.
Models were adapted from earlier work
by @cohen1995unexpected, excluding only two : Varley model (this is not a continuous one) 
and Malthus-Condorcet-Mill model (terms containing logarithm of extinct population does not make sense). 
All models assumed a generalized form : $\vec{P}_{t+1} = D \lambda_t \vec{P_t}$,
where $\vec{P_t}$ was a vector of population densities, 
the $i^{th}$ component $P_t^i$ representing
the density in the $i^{th}$ habitat patch ($i=1,2,\ldots,n$) at time $t$. 
The $n \times n$ matrix $\lambda(t)$ was diagonal with $i^{th}$ diagonal entry 
of stochastic density dependence form $\lambda_i(t)$ (see second column 
of Table \ref{tab_model_summary}). For simplicity, 
we considered $n$ patches in each model were arranged, evenly spaced, in a line
with the provision of "local" or "global" dispersal among patches. A n $\times$ n dispersal
matrix, $D$ assumed that a fraction $d$ of 
each population patch dispersed during any given time step, equally distributed to
the two or one nearest neighbors in case of local dispersal. For global dispersal, the fraction
$d$ that dispersed from each patch was divided up equally among all other patches. We called $d$ as dispersal
rate with $d$ being zero there was no dispersal, and threby no chance to rescue an extinct 
patch through migration from nearby metapopulation patches.

The environmental noises from any two out of $n$ patches, say, 
$(\epsilon_t^i, \epsilon_t^j)$ were normally 
distributed within each patch with mean $m=0$ and standard deviation $\sigma$ 
(see the specified value $\sigma$ for each model in 3$^{rd}$ 
column of Table \ref{tab_model_summary}). Noises were 
independent through time with same 
spatial correlations between patches (to within sampling error) for every simulation, 
but moderately (or extremely) left- and right-tail dependent between patches in
different simulations.

<!--Shya: ends here-->

A three dimensional array, $n_s \times n \times n_t$, 
$\epsilon$, was made from these extreme tail-dependence noise 
matrix generated via algorithm stated in 
Appendix \ref{SM-extr_tail_dep} and used to calculate extinction risk 
for each of the specified model in Table \ref{tab_model_summary} for $numsteps = 25$ years, $n =$ 5 patches
and $numsims = 10000$ simulations.
